---
title: "Import HoverNet and ProvGigaPath features with imageFeatureTCGA"
author: "Ilaria Billato"
date: "`r BiocStyle::doc_date()`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: imageFeatureTCGA
vignette: >
  %\VignetteIndexEntry{imageFeatureTCGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# imageFeatureTCGA

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
```

```{r load_packages, include=TRUE, results="hide", message=FALSE, warning=FALSE}
library(imageFeatureTCGA)
library(dplyr)
```

# Overview

`imageFeatureTCGA` (`imageTCGA`) provides convenient access to
histopathology-derived data from **TCGA** through two complementary pipelines:

- **HoVerNet** → cell segmentation and classification
- **ProvGigaPath** → slide- and tile-level embeddings


These datasets can be imported directly into R as **Bioconductor objects**,
facilitating downstream integration with TCGA omics and clinical data.


# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("waldronlab/imageFeatureTCGA")
```


# Available Data

Use the following function to download the catalog of available files:

```{r getcatalog}
getCatalog()
```

## Formats

- **HoVerNet** data is available in `JSON`, `GeoJSON`, `thumb` and `H5AD`
    formats.
- **ProvGigaPath** data is available in CSV format.


Note that the `thumb` format refers to the png thumbnails of the whole-slide
images.

### HoVerNet data


```{r listhover}
getCatalog("hovernet")
```


### ProvGigaPath data


```{r listprovgiga}
getCatalog("provgigapath")
```


# Importing HoVerNet data


You can import HoVerNet segmentation results as a
`SpatialExperiment` or `SpatialFeatureExperiment`.


```{r importHover}
getCatalog("hovernet") |>
    dplyr::filter(
        filename == paste(
            "TCGA-VG-A8LO-01A-01-DX1",
            "B39A4D64-82A1-4A04-8AB6-918F3058B83B",
            "json",
            "gz",
            sep = "."
        )
    ) |>
    getFileURLs() |>
    HoverNet(outClass = "SpatialExperiment") |>
    import()
```

Each cell is represented with:

- `x`, `y` spatial coordinates
- cell type and type probabilities
- optional contours stored in metadata


# Importing ProvGigaPath embeddings
80 

ProvGigaPath embeddings summarize tile or slide-level image features.


```{r importProvGiga}
getCatalog("provgigapath") |>
    dplyr::filter(
        filename == paste(
            "TCGA-VG-A8LO-01A-01-DX1",
            "B39A4D64-82A1-4A04-8AB6-918F3058B83B",
            "csv",
            "gz",
            sep = "."
        ) &
        level == "slide_level"
    ) |>
    getFileURLs() |>
    ProvGiga() |>
    import()
```


Each row corresponds to a slide, with an embedding vector describing the
image-derived features.


# See also

You can explore the full documentation through the package vignettes:

-[MOFA](https://github.com/waldronlab/imageFeatureTCGA/blob/devel/vignettes/MOFA_analysis.qmd)
-[Point Pattern Analysis](https://github.com/waldronlab/imageFeatureTCGA/blob/devel/vignettes/PPA.Rmd)

Note. More vignettes will be added as new feature types and workflows become
available.

# Shiny App: *imageTCGA*

The [imageTCGA](https://github.com/billila/imageTCGA) Shiny application
provides an interactive interface for exploring TCGA Diagnostic Image Database
metadata.

Click here to explore the shiny app:
[imageTCGA](https://shiny.sph.cuny.edu/app/imageTCGA/)

# Session Info

<details>
    <summary>Click here for Session Info</summary>
```{r sessioninfo}
sessionInfo()
```
</details>
