---
title: "Import Provgiga tile level embeddings and summary"
author: "Eslam Abousamra"
date: "2025-11-13"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: ImageFeatureTCGA
vignette: >
  %\VignetteIndexEntry{MOFA_COAD_READ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE}
library(ImageFeatureTCGA)
library(ggplot2)
library(dplyr)
library(sfdep)
library(spdep)
library(SpatialExperiment)

```


# Introduction

In this vignette, we demonstrate how to work with tile-level embeddings derived from whole-slide images (WSIs) of ovarian cancer. 
Each tile corresponds to a patch of the tissue, and its embedding is a high-dimensional vector capturing visual and morphological features extracted from provgigapath.


We begin by importing the tile-level data directly from the Cancer Data Science ProvGiga repository. 
Next, we perform principal component analysis (PCA) to reduce the high-dimensional embeddings to two principal components, which facilitates visualization and preliminary exploration of the data. 
We then visualize the spatial layout of the tiles on the tissue slide, coloring by the principal components to examine patterns in the embedding space. This workflow lays the groundwork for future spatial statistical analyses.





# Import tile level Embeddings. 

```{r}
import_tile_embedding <- function(slide_filenames, cancer = "TCGA_OV") {
  purrr::map(
    slide_filenames,
    ~ {
      url <- paste0(
        "https://store.cancerdatasci.org/provgigapath/tile_level/",
        cancer, "/", .x
      )
      df <- readr::read_csv(url, show_col_types = FALSE)
      df$slide_id <- .x
      df$source_file <- .x
      return(df)
    }
  ) |> setNames(slide_filenames)
}

```



```{r}

# list slides
ov_tile_files <- listProvGiga("TCGA_OV", level = "tile_level")$Filename

# select 2 slides as an example
ov_slides <- ov_tile_files[3]

# 3. Run the final import function
OV_tile_dfs <- import_tile_embedding(
  ov_slides,
  cancer = "TCGA_OV"
)


```

# Embedding PCA
```{r run PCA}
example_slide = OV_tile_dfs$`TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.csv.gz`
# Extract embedding numbers for pca
embedding_cols <- grep("^[0-9]+$", names(example_slide), value = TRUE)
# Run PCA on each tile
pca_res <- prcomp(example_slide[, embedding_cols], scale. = TRUE)
# create pca df
pca_example_slide <- bind_cols(
  example_slide,
  as_tibble(pca_res$x)[, 1:2] |> rename(PC1 = PC1, PC2 = PC2)
)

```


```{r visualize PC}


ggplot(pca_example_slide, aes(PC1, PC2)) +
  geom_point(alpha = 0.6, size = 1) +
  theme_minimal() +
  labs(title = "Tile-level PCA Ovarian Cancer Embedding: Single Slide")


ggplot(pca_example_slide, aes(tile_x, tile_y, color = PC1)) +
  geom_point(size = 1) +
  scale_color_viridis_c() +
  coord_equal() +
  theme_minimal() +
  labs(title = "Tissue layout colored by PC1")

```

# Spatial Patterns

To investigate spatial patterns in the tissue, we use the PCA-reduced embeddings for each tile. Each tile has a physical location (tile_x, tile_y) on the slide, which allows us to explore how similar embedding values cluster across space. We construct a k-nearest neighbor graph to define which tiles are spatially “connected,” and then compute global and local spatial autocorrelation metrics.


```{r}
coords <- pca_example_slide[, c("tile_x", "tile_y")]
nb <- knn2nb(knearneigh(coords, k = 6))
lw <- nb2listw(nb, style = "W")

```



Next, we calculate global spatial autocorrelation using Moran’s I and Geary’s C, which quantify the overall tendency of similar PC1 values to cluster or disperse on the tissue slide. We also compute Local Moran’s I (LISA) to detect local clusters of similar embedding values.

```{r}
mi <- moran.test(pca_example_slide$PC1, lw)
gc <- geary.test(pca_example_slide$PC1, lw)
lisa <- localmoran(pca_example_slide$PC1, lw)
pca_example_slide$localI <- lisa[, "Ii"]
pca_example_slide$localI_pval <- lisa[, "Pr(z != E(Ii))"]

print(mi)
print(gc)
```

We visualize the spatial patterns. The Moran scatterplot shows the relationship between each tile’s PC1 value and the mean of its neighbors, while the LISA plot highlights local clusters (“hotspots”) of high or low PC1 values across the tissue slide.

```{r}
moran.plot(pca_example_slide$PC1, lw, labels = FALSE,
           main = "Moran scatterplot of PC1")

# 5. LISA visualization
df_lisa <- data.frame(coords, Ii = lisa[, "Ii"])
ggplot(df_lisa, aes(x = tile_x, y = tile_y, color = Ii)) +
  geom_point(size = 0.5) +
  scale_color_viridis_c() +
  coord_equal() +
  theme_minimal() +
  ggtitle("Local Moran's I (LISA) for PC1")
```



# Adding HoverNet Nuclei Features

You can import HoVerNet segmentation results as a
`SpatialExperiment` or `SpatialFeatureExperiment`.


```{r importHover}
hov_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/TCGA_OV/json/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.json.gz"
)

hn_spe <- HoverNet(hov_file, outClass = "SpatialExperiment") |>
    import()

```

Each cell is represented with:

- `x`, `y` spatial coordinates
- cell type and type probabilities
- optional contours stored in metadata


```{r}
# Extract cell coordinates from HoverNet
cell_coords <- spatialCoords(hn_spe)

# Extract nuclei metadata (types, intensity, etc.)
cell_meta <- colData(hn_spe)


range(cell_meta$x); range(cell_meta$y)
range(pca_example_slide$tile_x); range(pca_example_slide$tile_y)
```


```{r}
plot(cell_meta$x, cell_meta$y, pch=16, col="#0000FF20")
points(pca_example_slide$tile_x, pca_example_slide$tile_y, pch=16, col="#FF000020")

```


# Scale factor between nuclei coordinates and tile coordinates
```{r}
sx <- diff(range(nuclei$x)) / diff(range(pca_example_slide$tile_x))
sy <- diff(range(nuclei$y)) / diff(range(pca_example_slide$tile_y))
sx; sy
scale_factor <- mean(c(sx, sy))
scale_factor
```



