---
title: "Import Provgiga tile level embeddings and summary"
author: "Eslam Abousamra"
date: "2025-11-13"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: ImageFeatureTCGA
vignette: >
  %\VignetteIndexEntry{MOFA_COAD_READ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r library import, message=FALSE}
library(BiocStyle)
library(ImageFeatureTCGA)
library(ggplot2)
library(dplyr)
library(sfdep)
library(spdep)
library(SpatialExperiment)
library(data.table)
```


# Introduction

In this vignette, we demonstrate how to work with tile-level embeddings derived 
from whole-slide images (WSIs) of ovarian cancer. 
Each tile corresponds to a patch of the tissue, and its embedding is a
high-dimensional vector capturing visual and morphological features 
extracted from provgigapath.


We begin by importing the tile-level data directly from the Cancer 
Data Science ProvGiga repository. 
Next, we perform principal component analysis (PCA) to reduce 
the high-dimensional embeddings to two principal components, 
which facilitates visualization and preliminary exploration of the data. 
We then visualize the spatial layout of the tiles on the tissue slide, 
coloring by the principal components to examine patterns in the embedding space.
This workflow lays the groundwork for future spatial statistical analyses.





# Import tile level Embeddings. 

```{r import tile-level emb, results = 'hide'}
# list slides
all_tile_files <- listProvGiga( level = "tile_level")[["Filename"]]

## filter with catalog
getCatalog("provgigapath") |> 
    dplyr::filter(Project.ID == "TCGA-OV") |> 
    dplyr::pull(filename)

# select Ovarian Cancer Slide as an example
tile_prov_url <- paste0(
    "https://store.cancerdatasci.org/provgigapath/tile_level/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.csv.gz"
)

example_slide <- ProvGiga(tile_prov_url) |>
    import()

```

# Embedding PCA
```{r run PCA}

# Extract embedding numbers for pca
embedding_cols <- grep("^[0-9]+$", names(example_slide), value = TRUE)
# Run PCA on each tile
pca_res <- prcomp(example_slide[, embedding_cols], scale. = TRUE)
# create pca df
pca_example_slide <- bind_cols(
    example_slide,
    as_tibble(pca_res$x)[, 1:2] |> rename(PC1 = "PC1", PC2 = "PC2")
)

```


```{r visualize PC}


ggplot(pca_example_slide, aes(PC1, PC2)) +
    geom_point(alpha = 0.6, size = 1) +
    theme_minimal() +
    labs(title = "Tile-level PCA Ovarian Cancer Embedding: Single Slide")


ggplot(pca_example_slide, aes(tile_x, tile_y, color = PC1)) +
    geom_point(size = 1) +
    scale_color_viridis_c() +
    coord_equal() +
    theme_minimal() +
    labs(title = "Tissue layout colored by PC1")

```

# Spatial Patterns

To investigate spatial patterns in the tissue, we use the PCA-reduced embeddings
for each tile. Each tile has a physical location (tile_x, tile_y) on the slide,
which allows us to explore how similar embedding values cluster across space. 
We construct a k-nearest neighbor graph to define which tiles are spatially 
“connected,” and then compute global and local spatial autocorrelation metrics.


```{r extract coords}
coords <- pca_example_slide[, c("tile_x", "tile_y")]
nb <- knn2nb(knearneigh(coords, k = 6))
lw <- nb2listw(nb, style = "W")

```



Next, we calculate global spatial autocorrelation using Moran’s I and Geary’s C,
which quantify the overall tendency of similar PC1 values to cluster or disperse
on the tissue slide. We also compute Local Moran’s I (LISA) to detect local 
clusters of similar embedding values.

```{r metrics}
mi <- moran.test(pca_example_slide$PC1, lw)
gc <- geary.test(pca_example_slide$PC1, lw)
lisa <- localmoran(pca_example_slide$PC1, lw)
pca_example_slide$localI <- lisa[, "Ii"]
pca_example_slide$localI_pval <- lisa[, "Pr(z != E(Ii))"]

print(mi)
print(gc)
```

We visualize the spatial patterns. The Moran scatterplot shows the relationship 
between each tile’s PC1 value and the mean of its neighbors, while the LISA plot
highlights local clusters (“hotspots”) of high or low PC1 values across the 
tissue slide.

```{r metrics 2}
moran.plot(pca_example_slide$PC1, lw, labels = FALSE,
                main = "Moran scatterplot of PC1")

# 5. LISA visualization
df_lisa <- data.frame(coords, Ii = lisa[, "Ii"])
ggplot(df_lisa, aes(x = tile_x, y = tile_y, color = Ii)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    coord_equal() +
    theme_minimal() +
    ggtitle("Local Moran's I (LISA) for PC1")
```



# Adding HoverNet Nuclei Features

You can import HoVerNet segmentation results as a
`SpatialExperiment` or `SpatialFeatureExperiment`.


```{r importHover}
hov_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/h5ad/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.h5ad.gz"
)


hn_spe <- HoverNet(hov_file, outClass = "SpatialExperiment") |>
    import()

```

Each cell is represented with:

- `x`, `y` spatial coordinates
- cell type and type probabilities
- optional contours stored in metadata


```{r cell coordinates}
# Extract cell coordinates from HoverNet
cell_coords <- spatialCoords(hn_spe)

# Extract nuclei metadata (types, intensity, etc.)
cell_meta <- colData(hn_spe)

```

# Visualizing Hovernet nuclei vs tile coordinates

```{r plotting hovernet vs tiles PCA}
plot(cell_meta$x, cell_meta$y, pch=16, col="#0000FF20")
points(pca_example_slide$tile_x, 
        pca_example_slide$tile_y, 
        pch=16, 
        col="#FF000020")

```


# Scale factor between nuclei coordinates and tile coordinates
```{r compute scaling factor}
# Step 1: Compute scale factor
# The nuclei coordinates (cell_meta$x/y) and tile coordinates 
# (tile_x/y) may be on slightly
# different scales. We compute a scale factor to align nuclei with tiles.
compute_scale_factor <- function(cell_meta, tile_df,
                                    cell_x = "x", cell_y = "y",
                                    tile_x = "tile_x", tile_y = "tile_y") {
    # adding a check to ensure correct structure
    stopifnot(
    all(c(cell_x, cell_y) %in% names(cell_meta)),
    all(c(tile_x, tile_y) %in% names(tile_df))
    )
    # calculate X and Y scale factors
    sx <- diff(range(cell_meta[[cell_x]], na.rm = TRUE)) /
        diff(range(tile_df[[tile_x]], na.rm = TRUE))

    sy <- diff(range(cell_meta[[cell_y]], na.rm = TRUE)) /
        diff(range(tile_df[[tile_y]], na.rm = TRUE))
    # mean scale factor
    scale_factor <- mean(c(sx, sy), na.rm = TRUE)

    list(
    scale_factor = scale_factor,
    sx = sx,
    sy = sy
    )
}

sf <- compute_scale_factor(cell_meta, pca_example_slide)
sf$scale_factor

# Step 2: Scale nuclei coordinates
cell_meta_scaled <- as.data.frame(cell_meta) %>%
    mutate(
        x = x / sf$scale_factor,
        y = y / sf$scale_factor
    )

# Step 3: Prepare tiles and nuclei data.tables
tile_size <- 224  # pixels per tile

# Tiles table: add bounding box coordinates for each tile
dt_tiles <- as.data.table(
    pca_example_slide %>%
        mutate(
            xmin = tile_x,
            xmax = tile_x + tile_size,
            ymin = tile_y,
            ymax = tile_y + tile_size
    )
)

# Nuclei table: each nucleus as a point (range start = range end)
# Use the 'label' column for descriptive cell type names (inflamm, nec, etc)
dt_nuc <- data.table(
    x1 = cell_meta_scaled$x,
    x2 = cell_meta_scaled$x,
    y1 = cell_meta_scaled$y,
    y2 = cell_meta_scaled$y,
    cell_type = cell_meta_scaled$label
)

# Step 4: Assign nuclei to tiles
# Use a non-equi join to match nuclei to tiles where the nucleus point falls within the tile bounds
assigned <- dt_tiles[dt_nuc, 
                        on = .(xmin <= x1, xmax >= x1, ymin <= y1, ymax >= y1), 
                        nomatch = 0L]

# Step 5: Compute per-tile dominant cell type
# Count the number of each cell type per tile
tile_counts <- assigned[, .N, by = .(tile_id, cell_type)]
setnames(tile_counts, "cell_type", "cell_type_label")
# Select the most frequent cell type in each tile
tile_dominant <- tile_counts[tile_counts[, .I[which.max(N)], by = tile_id]$V1]
setnames(tile_dominant, "cell_type_label", "dominant_cell_type")


# Step 6: Merge hovernet cell type back to tiles
pca_example_slide_hover <- merge(pca_example_slide, 
                                  tile_counts, 
                                  by = "tile_id",
                                  all.x = TRUE)
pca_example_slide_hover <- merge(pca_example_slide_hover, 
                            tile_dominant[, .(tile_id, dominant_cell_type)], 
                            by = "tile_id", all.x = TRUE)

```


```{r visualizing tile level with hovernet}

# Step 7: Plot per-tile cell types and dominant cell type
ggplot(pca_example_slide_hover, aes(tile_x, 
                                    tile_y, 
                                    color = cell_type_label, 
                                    size = N)) +
    geom_point(alpha = 0.7) +
    coord_equal() +
    theme_minimal() +
    labs(title = "All HoverNet cell types per tile")

ggplot(pca_example_slide_hover, aes(tile_x, tile_y, 
                                    color = dominant_cell_type)) +
    geom_point(size = 2) +
    coord_equal() +
    theme_minimal() +
    labs(title = "Per-tile dominant HoverNet cell type")


```


# Session Info

```{r sessioninfo}
sessionInfo()
```



