---
title: "Exploring HoverNet Features using ImageTCGA"
author: "Ilaria Billato"
date: "2025-11-13"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: ImageFeatureTCGA
vignette: >
  %\VignetteIndexEntry{ImageFeatureTCGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# ImageFeatureTCGA

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
```

```{r load_packages, include=TRUE, results="hide", message=FALSE, warning=FALSE}
library(ImageFeatureTCGA)
library(dplyr)
library(SpatialExperiment)
library(ggplot2)
```

# Introduction

HoverNet is a deep learning model for simultaneous segmentation and classification 
of nuclei in multi-tissue histology images. This vignette demonstrates how to import 
HoverNet output files into Bioconductor's spatial data structures and create 
visualizations of the segmentation results.

The package supports two main file formats:

- **JSON files** (`.json` or `.json.gz`): Contains cell coordinates, types, and 
  optional contours
- **H5AD files** (`.h5ad`): AnnData format with additional computed features like 
  mean intensity and nearest neighbor distance

# Importing HoverNet JSON Data

## From Local Files

First, let's download an example HoverNet JSON file:

```{r download-json}
json_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/json/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.json.gz")
```

## Import the JSON file into a `SpatialExperiment` object from URLs (with Automatic Caching)
The file will be automatically cached using `BiocFileCache`:

```{r import-json-url}
hn_spe <- HoverNet(json_file, outClass = "SpatialExperiment") |>
    import()
```

## Including Cell Contours

To include cell contour information in the metadata:

```{r import-contours}
hn_spe_contours <- HoverNet(
    json_file, 
    contours = TRUE,
    outClass = "SpatialExperiment"
) |> import()

# Access contours
contours <- metadata(hn_spe_contours)$contours
head(contours[[1]])
```

# Exploring the Data Structure

## Spatial Coordinates

```{r spatial-coords}
head(spatialCoords(hn_spe))
```

## Cell Metadata

```{r coldata}
head(colData(hn_spe))
```

## Cell Type Distribution

```{r cell-type-dist}
table(colData(hn_spe)$label)
```

## Type Map and Colors

The type map contains the mapping between cell types and their RGB colors:

```{r type-map}
type_map <- metadata(hn_spe)$type_map
type_map
```


# Basic Visualizations

## Simple Scatter Plot

```{r basic-plot, fig.width=8, fig.height=6}
coords <- data.frame(spatialCoords(hn_spe), colData(hn_spe))

ggplot(coords, aes(x = x, y = y, color = label)) +
    geom_point(size = 0.5, alpha = 0.8) +
    scale_color_manual(values = setNames(
        rgb(type_map$R/255, type_map$G/255, type_map$B/255),
        type_map$label
    )) +
    coord_fixed() +
    theme_minimal() +
    labs(
        title = "HoverNet Cell Segmentation",
        x = "X coordinate",
        y = "Y coordinate",
        color = "Cell Type"
    )
```

## Cell Type Probabilities

Visualize the confidence of cell type predictions:

```{r type-prob, fig.width=8, fig.height=6}
ggplot(coords, aes(x = x, y = y, color = type_prob)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    coord_fixed() +
    theme_minimal() +
    labs(
        title = "Cell Type Prediction Confidence",
        x = "X coordinate",
        y = "Y coordinate",
        color = "Type Probability"
    )
```

# Advanced Visualization: Overlay with Original Image

The package provides a convenient function to overlay the segmentation on the 
original tissue thumbnail image.

## Basic Overlay

```{r overlay-basic, fig.width=12, fig.height=6}
json_file <- "https://store.cancerdatasci.org/hovernet/TCGA_OV/json/TCGA-VG-A8LO-01A-01-DX1.B39A4D64-82A1-4A04-8AB6-918F3058B83B.json.gz"

plotHoverNetOverlay(json_file)
```

## Customized Overlay

```{r overlay-custom, fig.width=12, fig.height=6}
plotHoverNetOverlay(
    json_file,
    title = "Ovarian Cancer Tissue - Cell Segmentation",
    point_size = 0.02,
    legend_point_size = 3
)
```

## Custom Color Palette

```{r overlay-colors, fig.width=12, fig.height=6}
custom_colors <- c(
    "nolabe" = "#808080",
    "neopla" = "#E31A1C",
    "inflam" = "#1F78B4",
    "connec" = "#33A02C",
    "necros" = "#FF7F00",
    "no-neo" = "#6A3D9A"
)

plotHoverNetOverlay(
    json_file,
    color_palette = custom_colors,
    title = "Custom Color Scheme"
)
```

# Importing H5AD Data

H5AD files contain additional computed features like mean intensity and nearest 
neighbor distance.

```{r import-h5ad}
h5ad_url <- paste0(
    "https://store.cancerdatasci.org/hovernet/TCGA_OV/h5ad/",
    "TCGA-VG-A8LO-01A-01-DX1.B39A4D64-82A1-4A04-8AB6-918F3058B83B.h5ad"
)

h5ad_file <- file.path(tempdir(), basename(h5ad_url))
download.file(h5ad_url, destfile = h5ad_file)

hn_h5ad <- HoverNet(h5ad_file, outClass = "SpatialExperiment") |>
    import()

hn_h5ad
```

## Visualizing Additional Features

```{r h5ad-features, fig.width=10, fig.height=8}
h5ad_coords <- data.frame(spatialCoords(hn_h5ad), colData(hn_h5ad))

p1 <- ggplot(h5ad_coords, aes(x = x_centroid, y = y_centroid, 
                               color = mean_intensity)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    coord_fixed() +
    theme_minimal() +
    labs(title = "Mean Intensity", color = "Intensity")

p2 <- ggplot(h5ad_coords, aes(x = x_centroid, y = y_centroid, 
                               color = nearest_neighbor_distance)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c(option = "plasma") +
    coord_fixed() +
    theme_minimal() +
    labs(title = "Nearest Neighbor Distance", color = "Distance")

cowplot::plot_grid(p1, p2, ncol = 2)
```




# Session Info

<details>
    <summary>Click here for Session Info</summary>
```{r sessioninfo}
sessionInfo()
```
</details>
