---
title: "Point Pattern Analysis"
author: "Ilaria Billato"
date: "2025-11-13"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: ImageFeatureTCGA
vignette: >
  %\VignetteIndexEntry{ImageFeatureTCGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setting,include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```


```{r import libraries, message=FALSE, warning=FALSE}
library(ImageFeatureTCGA)
library(sfdep)
library(spdep)
library(SpatialFeatureExperiment)
library(SpatialExperiment)
library(ggplot2)
```

# Introduction


Point pattern analysis provides a framework to investigate the spatial
distribution of cells within histopathology images. When HoVerNet output
is imported through `ImageFeatureTCGA`, each cell is represented with
spatial coordinates and associated morphological or intensity features.
These data allow the computation of global and local spatial autocorrelation
metrics, such as Moran's I, Local Moran's I (LISA), Geary's C,
and Local Spatial Heteroscedasticity (LOSH).


In this vignette, we illustrate how to perform spatial autocorrelation
analysis starting from a `SpatialFeatureExperiment` object created from
HoVerNet segmentation results.


# Importing HoVerNet Data as a SpatialFeatureExperiment
```{r import hov as sfe, echo=TRUE}
hov_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/h5ad/",
    "TCGA-VG-A8LO-01A-01-DX1.B39A4D64-82A1-4A04-8AB6-918F3058B83B.h5ad.gz"
)

sfe <- HoverNet(hov_file, outClass = "SpatialFeatureExperiment") |>
    import()


sfe
```

We extract coordinates and select two example features: mean intensity
and nearest-neighbor distance.
```{r coords}
coords <- spatialCoords(sfe)
x <- colData(sfe)$mean_intensity
y <- colData(sfe)$nearest_neighbor_distance  
```


# Moran's I (Global Spatial Autocorrelation)


Moran's I quantifies the degree to which similar values cluster spatially. 
A positive Moran's I indicates spatial clustering, while negative values 
indicate spatial dispersion.

```{r moran}
nb <- knn2nb(knearneigh(coords, k = 6))
lw <- nb2listw(nb, style = "W")
mi <- moran.test(x, lw)
print(mi)
```

```{r moran plot}
moran.plot(x, lw, labels = FALSE, main = "Moran scatterplot")
```

# Local Moran (LISA)

```{r LISA}
lisa <- localmoran(x, lw)
colData(sfe)$localI <- lisa[, "Ii"]
colData(sfe)$localI_pval <- lisa[, "Pr(z != E(Ii))"]

df_lisa <- data.frame(coords, Ii = lisa[, "Ii"])
moran <- ggplot(df_lisa, aes(x = x, y = y, color = Ii)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    theme_minimal() +
    ggtitle("Local Moran's I")

moran
```

# Geary’s C (Global Spatial Autocorrelation)

Geary’s C is another measure of global autocorrelation but is more sensitive 
to local differences.
```{r Geary}
gc <- geary.test(x, lw)
print(gc)

gc_mc <- geary.mc(x, lw, nsim = 999)
print(gc_mc)
```

# Local Spatial Heteroscedasticity (LOSH)


LOSH characterizes local variance in the spatial distribution of a variable.


```{r LOSH}
losh <- losh(x, nb, lw$weights)
colData(sfe)$losh <- losh$hi           # valore LOSH
colData(sfe)$losh_z <- losh$z_hi       # z-score
colData(sfe)$losh_var <- losh$var_hi   # varianza

losh_pvals <- 2 * (1 - pnorm(abs(losh$z_hi)))
colData(sfe)$losh_pval <- losh_pvals

print(paste("LOSH medio:", round(mean(losh$hi, na.rm = TRUE), 4)))
print(paste("LOSH z-score medio:", round(mean(losh$z_hi, na.rm = TRUE), 4)))
print(paste("LOSH significativi (p<0.05):", 
            sum(losh_pvals < 0.05, na.rm = TRUE)))

df_losh <- data.frame(
    x = coords[,1], 
    y = coords[,2], 
    losh = losh$hi, 
    z_score = losh$z_hi,
    pval = losh_pvals
)

p_losh <- ggplot(df_losh, aes(x = x, y = y, color = losh)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_color_viridis_c() +
    theme_minimal() +
    ggtitle("Local Spatial Heteroscedasticity (LOSH)") 

p_losh

p_losh_z <- ggplot(df_losh, aes(x = x, y = y, color = z_score)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_color_gradient2(low = "blue", 
                            mid = "white", 
                            high = "red", 
                            midpoint = 20) +
    theme_minimal() +
    ggtitle("LOSH Z-scores") 

p_losh_z
```


# Session Info

<details>
    <summary>Click here for Session Info</summary>
```{r sessioninfo}
sessionInfo()
```
</details>
